---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sgsR - structurally guided sampling using ALS metrics

<!-- badges: start -->
<!-- badges: end -->

`sgsR` is designed to implement structurally guided sampling approaches for enhanced forest inventories. The package was designed to function using rasterized airborne laser scanning (ALS; Lidar) metrics to allow for stratification of forested areas based on structure.

## Installation

You can install the released version of sgsR from [Github](https://github.com/tgoodbody/sgsR) with:

``` r
install.packages("devtools")
devtools::install_github("https://github.com/tgoodbody/sgsR")
library(sgsR)
```

## Implementation {.unnumbered}

* Describe package fundamentals - `vignette("sgsR")`

* Overview of sampling algorithms - `vignette("sampling")`

* Overview of stratification algorithms - `vignette("stratification")`

* Overview of calculate algorithms - `vignette("calculating")`

## Collaborators {.unnumbered}

We are thankful for continued collaboration with academic, private industry, and government institutions to help improve `sgsR`. Special thanks to to:

```{r,echo=FALSE, results = 'asis'}
library(knitr)
names <- c("Martin Queinnec", "Joanne C. White", "Piotr Tompalski", "Andrew T. Hudak", "Ruben Valbuena", "Antoine LeBoeuf", "Ian Sinclair", "Grant McCartney", "Jean-Francois Prieur", "Murray Woods")

aff <- c("University of British Columbia", "Canadian Forest Service", "Canadian Forest Service", "United States Forest Service", " Bangor University", "Ministère des Forêts, de la Faune et des Parcs", "Ministry of Northern Development, Mines, Natural Resources and Forestry", "Forsite Consulting", "Laurentian Forestry Centre", "Ontario Ministry of Natural Resources")

urls <- c("https://www.researchgate.net/profile/Martin-Queinnec",
          "https://scholar.google.ca/citations?user=bqjk4skAAAAJ&hl=en",
          "https://scholar.google.ca/citations?user=RtYdz0cAAAAJ&hl=en",
          "https://scholar.google.ca/citations?hl=en&user=bdn7YVoAAAAJ",
          "https://scholar.google.com/citations?user=Nx336TQAAAAJ&hl=en",
          "https://scholar.google.com/citations?user=wGsKOK8AAAAJ&hl=en",
          "http://www.infogo.gov.on.ca/infogo/home.html#empProfile/332620/en",
          "https://www.signalhire.com/profiles/grant-mccartney%27s-email/99719223",
          "https://scf.rncan.gc.ca/employees/read/jprieur",
          "https://www.researchgate.net/profile/Murray-Woods")

df <- data.frame(Collaborator = names, Affiliation = aff)


df$Collaborator <- paste0("[", df$Collaborator, "](", urls, ")")

kable(df)

```


## Funding {.unnumbered}
Development of sgsR was made possible thanks to the financial support of the [Canadian Wood Fibre Centre's Forest Innovation Program](https://www.nrcan.gc.ca/science-and-data/funding-partnerships/funding-opportunities/forest-sector-funding-programs/forest-innovation-program/13137).

