---
title: "sampling"
output: rmarkdown::html_vignette
description: >
  Learn how to use sample_* functions.
vignette: >
  %\VignetteIndexEntry{sampling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r,warning=F,message=F,echo=FALSE}
library(sgsR)
library(terra)

#--- Load mraster and access files ---#
r <- system.file("extdata", "wall_metrics.tif", package = "sgsR")

#--- load the mraster using the terra package ---#
mraster <- terra::rast(r)

a <- system.file("extdata", "roads.shp", package = "sgsR")

#--- load the access vector using the sf package ---#
access <- sf::st_read(a, quiet = TRUE)

#--- apply kmeans algorithm to metrics raster ---#
sraster <- strat_kmeans(mraster = mraster, 
                        nStrata = 4) # algorithm will plot output

#--- apply kmeans algorithm to metrics raster ---#
existing <- sample_srs(raster = mraster, # use sraster as input for sampling
                       nSamp = 200, # request 200 samples be taken
                       mindist = 100) # algorithm will plot output

```

Currently, there are 6 functions associated with the `sample` verb in the `sgsR` package:

*   `sample_srs()` - simple random sampling

*   `sample_systematic()`- systematic sampling in a grid or hexagon tessellation 

*   `sample_strat()` - stratified sampling within a `sraster`

*   `sample_clhs()` - latin hypercube sampling algorithm

*   `sample_balanced()` - see [BalancedSampling](http://www.antongrafstrom.se/balancedsampling/)

*   `sample_ahels()` - adapted hypercube evaluation of a legacy sample (ahels) algorithm

## Access {#access .unnumbered}

One key feature of using the `sample_*` functions is its ability to define `access` corridors. Users can supply a road access network (must be `sf` line objects) and define buffers around `access` where samples should be excluded and included. 

Relevant and applicable parameters when `access` is defined are:

* `buff_inner` - Inner buffer is a parameter that defines the distance from the `access` where samples cannot be taken (i.e. if you don't want samples within 50 m of your `access` layer set `buff_inner = 50`). 

* `buff_outer` - Outer buffer is a parameter that defines the maximum distance that the samples can be located from `access` (i.e. if you don't want samples more than 200 meters from your `access` layer set `buff_inner = 200`).

## `sample_srs` {#srs .unnumbered}
We have demonstrated a simple example of using the `sample_srs()` function in `vignette("sgsR")`. We will demonstrate additional examples below. 

The input required for `sample_srs()` is a `raster`. This means that `sraster` and `mraster` are supported for this function.

```{r,warning=F,message=F} 
#--- perform simple random sampling ---#
sample_srs(raster = sraster, # input sraster
           nSamp = 200, # number of desired samples
           plot = TRUE) # plot
```

```{r,warning=F,message=F} 
sample_srs(raster = mraster, # input mraster
           nSamp = 200, # number of desired samples
           access = access, # define access road network
           mindist = 200, # minimum distance samples must be apart from one another
           buff_inner = 50, # inner buffer - no samples within this distance from road
           buff_outer = 200, # outer buffer - no samples further than this distance from road
           plot = TRUE) # plot
```

```{r,warning=F,message=F} 
sample_srs(raster = sraster, # input
           nSamp = 200, # number of desired samples
           access = access, # define access road network
           buff_inner = 50, # inner buffer - no samples within this distance from road
           buff_outer = 200, # outer buffer - no samples further than this distance from road
           plot = TRUE, # plot
           filename = tempfile(fileext = ".shp")) # write output samples to file
```

## `sample_systematic` {#systematic .unnumbered}

The `sample_systematic()` function applies systematic sampling across an area, the `cellsize` parameter defines resolution of the tessellation. The tesselation shape can be modified using the `square` parameter. Assigning `TRUE` (default) to the `square` parameter results in regular grids and assigning `False` to the `square` parameter results in hexagonal grids. The location of samples can also be adjusted using the `locations` parameter, where `centers` takes the center, `corners` takes all corners, and `random` takes a random location within each tessellation.

```{r,warning=F,message=F} 
#--- perform grid sampling ---#
sample_systematic(raster = sraster, # input sraster
                  cellsize = 1000, # grid distance
                  plot = TRUE) # plot
```

```{r,warning=F,message=F} 
#--- perform grid sampling ---#
sample_systematic(raster = sraster, # input sraster
                  cellsize = 500, # grid distance
                  square = FALSE, # hexagonal tessellation
                  location = "random", # random sample within tessellation
                  plot = TRUE) # plot
```      

```{r,warning=F,message=F} 
sample_systematic(raster = sraster, # input sraster
            cellsize = 500, # grid distance
            access = access, # define access road network
            buff_inner = 50, # inner buffer - no samples within this distance from road
            buff_outer = 200, # outer buffer - no samples further than this distance from road
            square = FALSE, # hexagonal tessellation
            location = "corners", # take corners instead of centers
            plot = TRUE)
```
## `sample_strat` {#sstrat .unnumbered}

The `sample_strat()` function contains a hierarchical sampling algorithm, which was originally developed by Martin Queinnec.

_Queinnec, M., White, J. C., & Coops, N. C. (2021). Comparing airborne and spaceborne photon-counting LiDAR canopy structural estimates across different boreal forest types. Remote Sensing of Environment, 262(August 2020), 112510._

This algorithm uses moving window (`wrow` and `wcol` parameters) to filter the input `sraster` to obtain locations where the stratum pixels are spatially grouped, rather than the dispersed individuals across the landscape.

The sampling is performed in 2 stages:

* **Rule 1** - Sample within spatially grouped stratum pixels. Moving window defined by `wrow` and `wcol`.

* **Rule 2** - If no more samples exist to satisfy desired sampling count, individual stratum pixels are sampled.

The rule applied to a select a particular sample is defined in the `rule` attribute of output samples. We give a few examples below:

```{r,warning=F,message=F} 
#--- perform stratified sampling random sampling ---#
sample_strat(sraster = sraster, # input sraster
             nSamp = 200, # desired sample number
             plot = TRUE) # plot
```

In some cases, users might want to include `existing` sample dataset within the algorithm. In order to adjust the total number of samples needed per stratum to reflect those already present in `existing`, we can use the utility function `extract_strata()`.

This function takes the input `sraster` and the `existing` sample dataset and extracts the stratum for each sample. These samples can inputted into the `sample_strat()` function which adjusts total samples required per class based on representation in `existing`.

```{r,warning=F,message=F} 
#--- extract strata values to existing samples ---#              
e.sr <- extract_strata(sraster = sraster, # input sraster
                       existing = existing) # existing samples to add strata value to

e.sr
```

Notice that `e.sr` now has an attribute named strata. If that parameter is not there, `sample_strat()` will give an error.

```{r,warning=F,message=F} 
sample_strat(sraster = sraster, # input sraster
             nSamp = 200, # desired sample number
             access = access, # define access road network
             existing = e.sr, # existing samples with strata values
             mindist = 200, # minimum distance samples must be apart from one another
             buff_inner = 50, # inner buffer - no samples within this distance from road
             buff_outer = 200, # outer buffer - no samples further than this distance from road
             plot = TRUE) # plot
```

As seen on the code in the example above, the defined `mindist` parameter specifies the minimum euclidean distance that samples must be apart from one another.

Notice that the sample outputs have `type` and `rule` attributes which outline whether the samples are `existing` or `new` and whether `rule1` or `rule2` were used to select the individual samples.

```{r,warning=F,message=F} 
sample_strat(sraster = sraster, # input
             nSamp = 200, # desired sample number
             access = access, # define access road network
             existing = e.sr, # existing samples with strata values
             include = TRUE, # include existing plots in nSamp total
             buff_inner = 50, # inner buffer - no samples within this distance from road
             buff_outer = 200, # outer buffer - no samples further than this distance from road
             filename = tempfile(fileext = ".shp"), # write output samples to file
             plot = TRUE) # plot
```

`include` parameter determines whether the existing samples should be included in the total count of samples defined by `nSamp`. By default, the `include` parameter is set as `FALSE`.

## `sample_clhs` {#clhs .unnumbered}

`sample_clhs()` function implements conditioned latin hypercube (clhs) sampling functionality.  A number of other functions in the `sgsR` package help to provide guidance on clhs sampling including `calculate_pop()` and `calculate_lhsOpt()`. Check out these functions to better understand how the sample numbers can be optimized.

The Syntax for this function is similar to others shown above, although parameters like `iter`, which define the number of iterations within the Metropolis-Hastings process are important to consider. In these examples we use a low `iter` value because it takes less time to run. Default values for `iter` within the `clhs` package are 10,000.

```{r,eval = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 200, # desired sample number
            plot = TRUE, # plot 
            iter = 100) # number of iterations
```

```{r,warning=F,message=F,echo=F,results = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 200, # desired sample number
            plot = TRUE, # plot 
            iter = 100) # number of iterations
```

```{r,eval = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 300, # desired sample number
            existing = existing, # existing samples
            iter = 100, # number of iterations
            details = TRUE, # output details
            plot = TRUE) # clhs details
```

```{r,warning=F,message=F,echo=F,results = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 300, # desired sample number
            existing = existing, # existing samples
            iter = 100, # number of iterations
            details = TRUE, # output details
            plot = TRUE) # clhs details
```

```{r,eval = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 300, # desired sample number
            iter = 100, # number of iterations
            existing = existing, # existing samples
            access = access, # define access road network
            buff_inner = 100, # inner buffer - no samples within this distance from road
            buff_outer = 300, # outer buffer - no samples further than this distance from road
            plot = TRUE) # plot
```

```{r,warning=F,message=F,echo=F,results = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 300, # desired sample number
            iter = 100, # number of iterations
            existing = existing, # existing samples
            access = access, # define access road network
            buff_inner = 100, # inner buffer - no samples within this distance from road
            buff_outer = 300, # outer buffer - no samples further than this distance from road
            plot = TRUE) # plot
```

`cost` parameter defines the `mraster` covariate, which is used to constrain the clhs sampling. This could be any number of variables. An example could be the distance a pixel is from road access (see example below), terrain slope, the output from `calculate_coobs()`, or many others.

```{r,warning=F,message=F} 
#--- cost constrained examples ---#
#--- calculate distance to access layer for each pixel in mr ---#
mr.c <- calculate_distance(raster = mraster, # input
                           access = access,
                           plot = TRUE) # define access road network

```

```{r,eval=F} 
sample_clhs(mraster = mr.c, # input
            nSamp = 250, # desired sample number
            iter = 100, # number of iterations
            cost = "dist2access", # cost parameter - name defined in calculate_distance()
            plot = TRUE) # plot
```

```{r,warning=F,message=F,echo=F,results = FALSE} 
sample_clhs(mraster = mr.c, # input
            nSamp = 250, # desired sample number
            iter = 100, # number of iterations
            cost = "dist2access", # cost parameter - name defined in calculate_distance()
            plot = TRUE) # plot
```

```{r,eval = FALSE} 
sample_clhs(mraster = mr.c, # input
            nSamp = 250, # desired sample number
            existing = existing, # existing samples
            iter = 100, # number of iterations
            cost = "dist2access", # cost parameter - name defined in calculate_distance()
            plot = TRUE) # plot

```

```{r,warning=F,message=F,echo=F,results = FALSE} 
sample_clhs(mraster = mr.c, # input
            nSamp = 250, # desired sample number
            existing = existing, # existing samples
            iter = 100, # number of iterations
            cost = "dist2access", # cost parameter - name defined in calculate_distance()
            plot = TRUE) # plot

```

## `sample_balanced` {#balanced .unnumbered}

The `sample_balanced()` algorithm performs a balanced sampling methodology from the [`stratifyR / SamplingBigData`](http://www.antongrafstrom.se/balancedsampling/) packages.

```{r,warning=F,message=F} 
sample_balanced(mraster = mraster, # input
                nSamp = 200, # desired sample number
                plot = TRUE) # plot
```

```{r,warning=F,message=F} 
sample_balanced(mraster = mraster, # input
                nSamp = 100, # desired sample number
                algorithm = "lcube", # algorithm type
                access = access, # define access road network
                buff_inner = 50, # inner buffer - no samples within this distance from road
                buff_outer = 200) # outer buffer - no samples further than this distance from road
```

## `sample_ahels` {#ahels .unnumbered}

The `sample_ahels()` function performs the adapted Hypercube Evaluation of a Legacy Sample (ahels) algorithm using`existing` sample data and an `mraster`. New samples are allocated based on quantile ratios between the `existing` sample and `mraster` covariate dataset.

This algorithm:

1. Determines the quantile distributions of `existing` samples and `mraster` covariates.

2. Determines quantiles where there is a disparity between samples and covariates.

3. Prioritizes sampling within those quantile to improve representation.

To use this function, user must first specify the number of quantiles (`nQuant`) followed by either the `nSamp` (total number of desired samples to be added) or the `threshold` (proportional representation between the sample and the ecovariate quantiles - default is 0.9) parameters. We recommended you setting the `threshold` values at or below 0.9 as higher values might cause the algorithm to add samples repeatedly.

```{r,eval = FALSE}
sample_ahels(mraster = mraster[[1:3]], # input mraster - first 3 layers only
             existing = existing, # existing samples
             plot = TRUE) # plot
```

```{r,warning=F,message=F,echo=FALSE, results = FALSE}
s <- sample_ahels(mraster = mraster[[1:3]], # input mraster - first 3 layers only
             existing = existing, # existing samples
             plot = TRUE) # plot
```
```{r,echo=FALSE}
s
```

Notice that no `threshold`, `nSamp`, or `nQuant` were defined. That is because the default setting for `threshold = 0.9` and `nQuant = 10`.

The first matrix output shows the quantile ratios between the sample and the covariates. A value of 1.0 indicates that the samples and the covariates are equally represented. Values above 1.0 indicates over representation of the sample, while less than the value 1.0 indicates under representation of the sample.

```{r,eval = FALSE}
sample_ahels(mraster = mraster[[1:3]], # input mraster - first 3 layers only
             existing = existing, # existing samples
             nQuant = 20, # define 20 quantiles
             nSamp = 300, # total samples desired
             filename = tempfile(fileext = ".shp")) # write samples to disc
```

```{r,warning=F,message=F,echo=FALSE, results = FALSE}
s <- sample_ahels(mraster = mraster[[1:3]], # input mraster - first 3 layers only
             existing = existing, # existing samples
             nQuant = 20, # define 20 quantiles
             nSamp = 300, # total samples desired
             plot = TRUE,
             filename = tempfile(fileext = ".shp")) # write samples to disc

```

```{r,echo=FALSE}
s
```
Notice that the total number of samples is 500. This value is the sum of existing samples (200) and number of samples defined by `nSamp = 300`.
