---
title: "sampling"
output: rmarkdown::html_vignette
description: >
  Learn how to use sample_* functions.
vignette: >
  %\VignetteIndexEntry{sampling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r,warning=F,message=F,echo=FALSE}
library(sgsR)
library(terra)

#--- Load mraster and access files ---#
r <- system.file("extdata", "mraster.tif", package = "sgsR")

#--- load the mraster using the terra package ---#
mraster <- terra::rast(r)

a <- system.file("extdata", "access.shp", package = "sgsR")

#--- load the access vector using the sf package ---#
access <- sf::st_read(a, quiet = TRUE)

#--- apply quantiles algorithm to metrics raster ---#
sraster <- strat_quantiles(mraster = mraster$zq90, # use mraster as input for sampling
                           nStrata = 4) # algorithm will produce 4 strata

#--- apply stratified sampling algorithm ---#
existing <- sample_strat(sraster = sraster, # use mraster as input for sampling
                         nSamp = 200, # request 200 samples be taken
                         mindist = 100) # define that samples must be 100 m apart

```

Currently, there are 6 functions associated with the `sample` verb in the `sgsR` package:

*   `sample_srs()` - simple random sampling

*   `sample_systematic()`- systematic sampling in a grid or hexagon tessellation 

*   `sample_strat()` - stratified sampling within a `sraster`

*   `sample_clhs()` - Latin hypercube sampling

*   `sample_balanced()` - see [BalancedSampling](http://www.antongrafstrom.se/balancedsampling/)

*   `sample_ahels()` - adapted hypercube evaluation of a legacy sample (ahels)

## Access {#access .unnumbered}

One key feature of using some `sample_*` functions is its ability to define `access` corridors. Users can supply a road `access` network (must be `sf` line objects) and define buffers around `access` where samples should be excluded and included. 

Relevant and applicable parameters when `access` is defined are:

* `buff_inner` - Inner buffer parameter that defines the distance from `access` where samples cannot be taken (i.e. if you don't want samples within 50 m of your `access` layer set `buff_inner = 50`). 

* `buff_outer` - Outer buffer parameter that defines the maximum distance that the samples can be located from `access` (i.e. if you don't want samples more than 200 meters from your `access` layer set `buff_inner = 200`).

## `sample_srs` {#srs .unnumbered}
We have demonstrated a simple example of using the `sample_srs()` function in `vignette("sgsR")`. We will demonstrate additional examples below. 

The input required for `sample_srs()` is a `raster`. This means that `sraster` and `mraster` are supported for this function.

```{r,warning=F,message=F} 
#--- perform simple random sampling ---#
sample_srs(raster = sraster, # input sraster
           nSamp = 200, # number of desired samples
           plot = TRUE) # plot
```

```{r,warning=F,message=F} 
sample_srs(raster = mraster, # input mraster
           nSamp = 200, # number of desired samples
           access = access, # define access road network
           mindist = 200, # minimum distance samples must be apart from one another
           buff_inner = 50, # inner buffer - no samples within this distance from road
           buff_outer = 200, # outer buffer - no samples further than this distance from road
           plot = TRUE) # plot
```

## `sample_systematic` {#systematic .unnumbered}

The `sample_systematic()` function applies systematic sampling across an area with the `cellsize` parameter defining the resolution of the tessellation. The tessellation shape can be modified using the `square` parameter. Assigning `TRUE` (default) to the `square` parameter results in a regular grid and assigning `FALSE` results in a hexagonal grid. The location of samples can also be adjusted using the `locations` parameter, where `centers` takes the center, `corners` takes all corners, and `random` takes a random location within each tessellation.

```{r,warning=F,message=F} 
#--- perform grid sampling ---#
sample_systematic(raster = sraster, # input sraster
                  cellsize = 1000, # grid distance
                  plot = TRUE) # plot
```

```{r,warning=F,message=F} 
#--- perform grid sampling ---#
sample_systematic(raster = sraster, # input sraster
                  cellsize = 500, # grid distance
                  square = FALSE, # hexagonal tessellation
                  location = "random", # random sample within tessellation
                  plot = TRUE) # plot
```      

```{r,warning=F,message=F} 
sample_systematic(raster = sraster, # input sraster
            cellsize = 500, # grid distance
            access = access, # define access road network
            buff_inner = 50, # inner buffer - no samples within this distance from road
            buff_outer = 200, # outer buffer - no samples further than this distance from road
            square = FALSE, # hexagonal tessellation
            location = "corners", # take corners instead of centers
            plot = TRUE)
```

## `sample_strat` {#sstrat .unnumbered}

The `sample_strat()` function contains a hierarchical sampling algorithm, which was originally developed by Martin Queinnec.

_Queinnec, M., White, J. C., & Coops, N. C. (2021). Comparing airborne and spaceborne photon-counting LiDAR canopy structural estimates across different boreal forest types. Remote Sensing of Environment, 262(August 2020), 112510._

This algorithm uses moving window (`wrow` and `wcol` parameters) to filter the input `sraster` to prioritize sample locations where stratum pixels are spatially grouped, rather than dispersed individuals across the landscape.

Sampling is performed using 2 rules:

* **Rule 1** - Sample within spatially grouped stratum pixels. Moving window defined by `wrow` and `wcol`.

* **Rule 2** - If no more samples exist to satisfy desired sampling count, individual stratum pixels are sampled.

The rule applied to a select a particular sample is defined in the `rule` attribute of output samples. We give a few examples below:

```{r,warning=F,message=F} 
#--- perform stratified sampling random sampling ---#
sample_strat(sraster = sraster, # input sraster
             nSamp = 200, # desired sample number
             plot = TRUE) # plot
```

In some cases, users might want to include `existing` samples within the algorithm. In order to adjust the total number of samples needed per stratum to reflect those already present in `existing`, we can use the intermediate function `extract_strata()`.

This function uses the `sraster` and `existing` samples and extracts the stratum for each. These samples can be included within `sample_strat()`, which adjusts total samples required per class based on representation in `existing`.

```{r,warning=F,message=F} 
#--- extract strata values to existing samples ---#              
e.sr <- extract_strata(sraster = sraster, # input sraster
                       existing = existing) # existing samples to add strata value to

e.sr
```

Notice that `e.sr` now has an attribute named strata. If that parameter is not there, `sample_strat()` will give an error.

```{r,warning=F,message=F} 
sample_strat(sraster = sraster, # input sraster
             nSamp = 200, # desired sample number
             access = access, # define access road network
             existing = e.sr, # existing samples with strata values
             mindist = 200, # minimum distance samples must be apart from one another
             buff_inner = 50, # inner buffer - no samples within this distance from road
             buff_outer = 200, # outer buffer - no samples further than this distance from road
             plot = TRUE) # plot
```

As seen on the code in the example above, the defined `mindist` parameter specifies the minimum euclidean distance that samples must be apart from one another.

Notice that the sample outputs have `type` and `rule` attributes which outline whether the samples are `existing` or `new` and whether `rule1` or `rule2` were used to select the individual samples.

```{r,warning=F,message=F} 
sample_strat(sraster = sraster, # input
             nSamp = 200, # desired sample number
             access = access, # define access road network
             existing = e.sr, # existing samples with strata values
             include = TRUE, # include existing plots in nSamp total
             buff_inner = 50, # inner buffer - no samples within this distance from road
             buff_outer = 200, # outer buffer - no samples further than this distance from road
             filename = tempfile(fileext = ".shp"), # write output samples to file
             plot = TRUE) # plot
```

The `include` parameter determines whether `existing` samples should be included in the total count of samples defined by `nSamp`. By default, the `include` parameter is set as `FALSE`.

## `sample_clhs` {#clhs .unnumbered}

`sample_clhs()` function implements conditioned Latin hypercube (clhs) sampling methodology from the [`clhs`](https://cran.r-project.org/web/packages/clhs/clhs.pdf) package.  A number of other functions in the `sgsR` package help to provide guidance on clhs sampling including `calculate_pop()` and `calculate_lhsOpt()`. Check out these functions to better understand how sample numbers could be optimized.

The syntax for this function is similar to others shown above, although parameters like `iter`, which define the number of iterations within the Metropolis-Hastings process are important to consider. In these examples we use a low `iter` value because it takes less time to run. Default values for `iter` within the `clhs` package are 10,000.

```{r,eval = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 200, # desired sample number
            plot = TRUE, # plot 
            iter = 100) # number of iterations
```

```{r,warning=F,message=F,echo=F,results = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 200, # desired sample number
            plot = TRUE, # plot 
            iter = 100) # number of iterations
```

```{r,eval = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 300, # desired sample number
            iter = 100, # number of iterations
            existing = existing, # existing samples
            access = access, # define access road network
            buff_inner = 100, # inner buffer - no samples within this distance from road
            buff_outer = 300, # outer buffer - no samples further than this distance from road
            plot = TRUE) # plot
```

```{r,warning=F,message=F,echo=F,results = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 300, # desired sample number
            iter = 100, # number of iterations
            existing = existing, # existing samples
            access = access, # define access road network
            buff_inner = 100, # inner buffer - no samples within this distance from road
            buff_outer = 300, # outer buffer - no samples further than this distance from road
            plot = TRUE) # plot
```

The `cost` parameter defines the `mraster` covariate, which is used to constrain the clhs sampling. This could be any number of variables. An example could be the distance a pixel is from road `access` (e.g. from `calculate_distance()` see example below), terrain slope, the output from `calculate_coobs()`, or many others.

```{r,warning=F,message=F} 
#--- cost constrained examples ---#
#--- calculate distance to access layer for each pixel in mr ---#
mr.c <- calculate_distance(raster = mraster, # input
                           access = access,
                           plot = TRUE) # define access road network

```

```{r,eval=F} 
sample_clhs(mraster = mr.c, # input
            nSamp = 250, # desired sample number
            iter = 100, # number of iterations
            cost = "dist2access", # cost parameter - name defined in calculate_distance()
            plot = TRUE) # plot
```

```{r,warning=F,message=F,echo=F,results = FALSE} 
sample_clhs(mraster = mr.c, # input
            nSamp = 250, # desired sample number
            iter = 100, # number of iterations
            cost = "dist2access", # cost parameter - name defined in calculate_distance()
            plot = TRUE) # plot
```

## `sample_balanced` {#balanced .unnumbered}

The `sample_balanced()` algorithm performs a balanced sampling methodology from the [`stratifyR / SamplingBigData`](http://www.antongrafstrom.se/balancedsampling/) packages.

```{r,warning=F,message=F} 
sample_balanced(mraster = mraster, # input
                nSamp = 200, # desired sample number
                plot = TRUE) # plot
```

```{r,warning=F,message=F} 
sample_balanced(mraster = mraster, # input
                nSamp = 100, # desired sample number
                algorithm = "lcube", # algorithm type
                access = access, # define access road network
                buff_inner = 50, # inner buffer - no samples within this distance from road
                buff_outer = 200) # outer buffer - no samples further than this distance from road
```

## `sample_ahels` {#ahels .unnumbered}

The `sample_ahels()` function performs the adapted Hypercube Evaluation of a Legacy Sample (ahels) algorithm using`existing` sample data and an `mraster`. New samples are allocated based on quantile ratios between the `existing` sample and `mraster` covariate dataset.

This algorithm was adapted from that presented in the paper below, which we highly recommend.

_Malone BP, Minansy B, Brungard C. 2019. Some methods to improve the utility of conditioned Latin hypercube sampling. PeerJ 7:e6451 DOI 10.7717/peerj.6451_ 

This algorithm:

1. Determines the quantile distributions of `existing` samples and `mraster` covariates.

2. Determines quantiles where there is a disparity between samples and covariates.

3. Prioritizes sampling within those quantile to improve representation.

To use this function, user must first specify the number of quantiles (`nQuant`) followed by either the `nSamp` (total number of desired samples to be added) or the `threshold` (sampling ratio vs. covariate coverage ratio for quantiles - default is 0.9) parameters. We recommended you setting the `threshold` values at or below 0.9.

```{r,eval = FALSE}
sample_ahels(mraster = mraster, 
             existing = existing, # existing samples
             plot = TRUE) # plot
```

```{r,warning=F,message=F,echo=FALSE, results = FALSE}
s <- sample_ahels(mraster = mraster, 
             existing = existing, # existing samples
             plot = TRUE) # plot
```
```{r,echo=FALSE}
s
```

Notice that no `threshold`, `nSamp`, or `nQuant` were defined. That is because the default setting for `threshold = 0.9` and `nQuant = 10`.

The first matrix output shows the quantile ratios between the sample and the covariates. A value of 1.0 indicates that samples are represented relative to the quantile coverage. Values > 1.0 indicate over representation of samples, while < 1.0 indicate under representation of samples.

```{r,eval = FALSE}
sample_ahels(mraster = mraster, 
             existing = existing, # existing samples
             nQuant = 20, # define 20 quantiles
             nSamp = 300, # total samples desired
             filename = tempfile(fileext = ".shp")) # write samples to disc
```

```{r,warning=F,message=F,echo=FALSE, results = FALSE}
s <- sample_ahels(mraster = mraster, 
             existing = existing, # existing samples
             nQuant = 20, # define 20 quantiles
             nSamp = 300, # total samples desired
             plot = TRUE,
             filename = tempfile(fileext = ".shp")) # write samples to disc

```

```{r,echo=FALSE}
s
```

Notice that the total number of samples is 500. This value is the sum of existing samples (200) and number of samples defined by `nSamp = 300`.
