---
title: "sampling"
output: rmarkdown::html_vignette
description: >
  Learn how to use sample_* functions.
vignette: >
  %\VignetteIndexEntry{sampling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r,warning=F,message=F,echo=FALSE}
library(sgsR)
library(terra)

#--- Load mraster and access files ---#
r <- system.file("extdata", "wall_metrics.tif", package = "sgsR")

#--- load the mraster using the terra package ---#
mraster <- terra::rast(r)

a <- system.file("extdata", "roads.shp", package = "sgsR")

#--- load the access vector using the sf package ---#
access <- sf::st_read(a, quiet = TRUE)

#--- apply kmeans algorithm to metrics raster ---#
sraster <- strat_kmeans(mraster = mraster, 
                        nStrata = 4) # algorithm will plot output

#--- apply kmeans algorithm to metrics raster ---#
existing <- sample_srs(raster = mraster, # use sraster as input for sampling
                       nSamp = 200, # request 200 samples be taken
                       mindist = 100) # algorithm will plot output

```

A total of 5 `sample` functions currently exist in the package:

*   `sample_srs()` - simple random sampling

*   `sample_systematic()`- systematic sampling in a grid or hexagon tessellation 

*   `sample_strat()` - stratified sampling within a `sraster`

*   `sample_clhs()` - Latin hypercube sampling algorithm

*   `sample_balanced()` - see [BalancedSampling](http://www.antongrafstrom.se/balancedsampling/)

*   `sample_ahels()` - Adapted hypercube evaluation of a legacy sample (ahels) algorithm

## Access {#access .unnumbered}

A feature of the `sample_*` functions is the ability to define `access` corridors. Users can supply a road access network (must be `sf` line objects) and define buffers around `access` where samples should be excluded and included. 

Important additional parameters when `access` is provided are:

* `buff_inner` - An inner buffer that defines the distance from `access` where samples cannot be taken (i.e. if you don't want samples within 50 m of your `access` layer set `buff_inner = 50`). 

* `buff_outer` - Maximum distance samples can be located from `access` (i.e. if you don't want samples more than 200 meters from your `access` layer set `buff_inner = 200`)

## `sample_srs` {#srs .unnumbered}
We provided a very simple example of using the `sample_srs()` function in `vignette("sgsR")`. We provide some additional examples below. 

Notice that the input for `sample_srs()` is `raster`. This means that either `sraster` or an `mraster` are supported.

```{r,warning=F,message=F} 
#--- perform simple random sampling ---#
sample_srs(raster = sraster, # input sraster
           nSamp = 200, # number of desired samples
           plot = TRUE) # plot
```

```{r,warning=F,message=F} 
sample_srs(raster = mraster, # input mraster
           nSamp = 200, # number of desired samples
           access = access, # define access road network
           mindist = 200, # minimum distance samples must be apart from one another
           buff_inner = 50, # inner buffer - no samples within this distance from road
           buff_outer = 200, # outer buffer - no samples further than this distance from road
           plot = TRUE) # plot
```

```{r,warning=F,message=F} 
sample_srs(raster = sraster, # input
           nSamp = 200, # number of desired samples
           access = access, # define access road network
           buff_inner = 50, # inner buffer - no samples within this distance from road
           buff_outer = 200, # outer buffer - no samples further than this distance from road
           plot = TRUE, # plot
           filename = tempfile(fileext = ".shp")) # write output samples to file
```

## `sample_systematic` {#systematic .unnumbered}

The `sample_systematic()` function applies systematic sampling across an area where the `cellsize` parameter defines resolution of the tessellation. Tesselation shape options are defined by the `square` parameter, which is a regular grid when `TRUE` (default) and hexagonal when `FALSE`. The location of samples can also be adjusted using the `locations` parameter, where `centers` takes the center, `corners` takes all corners, and `random` takes a random location within each tessellation.

```{r,warning=F,message=F} 
#--- perform grid sampling ---#
sample_systematic(raster = sraster, # input sraster
                  cellsize = 1000, # grid distance
                  plot = TRUE) # plot
```

```{r,warning=F,message=F} 
#--- perform grid sampling ---#
sample_systematic(raster = sraster, # input sraster
                  cellsize = 500, # grid distance
                  square = FALSE, # hexagonal tessellation
                  location = "random", # random sample within tessellation
                  plot = TRUE) # plot
```      

```{r,warning=F,message=F} 
sample_systematic(raster = sraster, # input sraster
            cellsize = 500, # grid distance
            access = access, # define access road network
            buff_inner = 50, # inner buffer - no samples within this distance from road
            buff_outer = 200, # outer buffer - no samples further than this distance from road
            square = FALSE, # hexagonal tessellation
            location = "corners", # take corners instead of centers
            plot = TRUE)
```
## `sample_strat` {#sstrat .unnumbered}

The `sample_strat()` function contains a hierarchical sampling algorithm originally developed my Martin Queinnec.

_Queinnec, M., White, J. C., & Coops, N. C. (2021). Comparing airborne and spaceborne photon-counting LiDAR canopy structural estimates across different boreal forest types. Remote Sensing of Environment, 262(August 2020), 112510._

The algorithm uses a moving window (`wrow` and `wcol` parameters) to filter the input `sraster` for locations where stratum pixels are spatially grouped rather than dispersed individually across the landscape.

The sampling is performed in 2 stages:

* **Rule 1** - Sample within spatially grouped stratum pixels. Moving window defined by `wrow` and `wcol`.

* **Rule 2** - If no more samples exist to satisfy desired sampling count, individual stratum pixels are sampled.

The rule applied to a select a particular sample is defined in the `rule` attribute of output samples. We give a few examples below:

```{r,warning=F,message=F} 
#--- perform stratified sampling random sampling ---#
sample_strat(sraster = sraster, # input sraster
             nSamp = 200, # desired sample number
             plot = TRUE) # plot
```

In some cases, a user may wish to include an `existing` sample data set within the algorithm. In order to adjust the total number of samples needed per stratum to reflect those already present in `existing`, we can use the utility function `extract_strata()`.

This function takes an input `sraster` and the `existing` sample data set and extracts the stratum for each sample. These samples can then be input into the `sample_strat()` function which adjusts total required sample per class based on representation in `existing`.

```{r,warning=F,message=F} 
#--- extract strata values to existing samples ---#              
e.sr <- extract_strata(sraster = sraster, # input sraster
                       existing = existing) # existing samples to add strata value to

e.sr
```

Notice that `e.sr` now has an attribute named strata. If that parameter is not there, `sample_strat()` will give an error.

```{r,warning=F,message=F} 
sample_strat(sraster = sraster, # input sraster
             nSamp = 200, # desired sample number
             access = access, # define access road network
             existing = e.sr, # existing samples with strata values
             mindist = 200, # minimum distance samples must be apart from one another
             buff_inner = 50, # inner buffer - no samples within this distance from road
             buff_outer = 200, # outer buffer - no samples further than this distance from road
             plot = TRUE) # plot
```

The `mindist` parameter defined in the above example specifies the minimum euclidian distance that samples must be apat from one another.

Notice that the sample outputs have `type` and `rule` attributes which outline whether samples are `existing` or `new` and whether `rule1` or `rule2` were used to select individual samples.

```{r,warning=F,message=F} 
sample_strat(sraster = sraster, # input
             nSamp = 200, # desired sample number
             access = access, # define access road network
             existing = e.sr, # existing samples with strata values
             include = TRUE, # include existing plots in nSamp total
             buff_inner = 50, # inner buffer - no samples within this distance from road
             buff_outer = 200, # outer buffer - no samples further than this distance from road
             filename = tempfile(fileext = ".shp"), # write output samples to file
             plot = TRUE) # plot
```

The `include` parameter determines whether existing samples should be included in the total count of samples defined by `nSamp`. By defaults `include = FALSE`.

## `sample_clhs` {#clhs .unnumbered}

The `sample_clhs()` function implements conditioned latin hypercube (clhs) sampling functionality.  A number of other functions in the `sgsR` package help to provide guidance on clhs sampling including `calculate_pop()` and `calculate_lhsOpt()`. Be sure to check out those functions for better understanding of how to optimize sample numbers.

Syntax for this function is similar to others shown above, though parameters like `iter`, which define the number of iterations within the Metropolis-Hastings process are important to consider. In these examples we use a low `iter` value because it takes less time to run. Default values for `iter` within the `clhs` package are 10,000.

```{r,eval = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 200, # desired sample number
            plot = TRUE, # plot 
            iter = 100) # number of iterations
```

```{r,warning=F,message=F,echo=F,results = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 200, # desired sample number
            plot = TRUE, # plot 
            iter = 100) # number of iterations
```

```{r,eval = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 300, # desired sample number
            existing = existing, # existing samples
            iter = 100, # number of iterations
            details = TRUE, # output details
            plot = TRUE) # clhs details
```

```{r,warning=F,message=F,echo=F,results = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 300, # desired sample number
            existing = existing, # existing samples
            iter = 100, # number of iterations
            details = TRUE, # output details
            plot = TRUE) # clhs details
```

```{r,eval = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 300, # desired sample number
            iter = 100, # number of iterations
            existing = existing, # existing samples
            access = access, # define access road network
            buff_inner = 100, # inner buffer - no samples within this distance from road
            buff_outer = 300, # outer buffer - no samples further than this distance from road
            plot = TRUE) # plot
```

```{r,warning=F,message=F,echo=F,results = FALSE} 
sample_clhs(mraster = mraster, # input
            nSamp = 300, # desired sample number
            iter = 100, # number of iterations
            existing = existing, # existing samples
            access = access, # define access road network
            buff_inner = 100, # inner buffer - no samples within this distance from road
            buff_outer = 300, # outer buffer - no samples further than this distance from road
            plot = TRUE) # plot
```

The `cost` parameter defines an `mraster` covariate to be used to constrain the clhs sampling. This could be any number of variables. An example could be the distance a pixel is from road access (see example below), terrain slope, the output from `calculate_coobs()`, or many others.

```{r,warning=F,message=F} 
#--- cost constrained examples ---#
#--- calculate distance to access layer for each pixel in mr ---#
mr.c <- calculate_distance(raster = mraster, # input
                           access = access,
                           plot = TRUE) # define access road network

```

```{r,eval=F} 
sample_clhs(mraster = mr.c, # input
            nSamp = 250, # desired sample number
            iter = 100, # number of iterations
            cost = "dist2access", # cost parameter - name defined in calculate_distance()
            plot = TRUE) # plot
```

```{r,warning=F,message=F,echo=F,results = FALSE} 
sample_clhs(mraster = mr.c, # input
            nSamp = 250, # desired sample number
            iter = 100, # number of iterations
            cost = "dist2access", # cost parameter - name defined in calculate_distance()
            plot = TRUE) # plot
```

```{r,eval = FALSE} 
sample_clhs(mraster = mr.c, # input
            nSamp = 250, # desired sample number
            existing = existing, # existing samples
            iter = 100, # number of iterations
            cost = "dist2access", # cost parameter - name defined in calculate_distance()
            plot = TRUE) # plot

```

```{r,warning=F,message=F,echo=F,results = FALSE} 
sample_clhs(mraster = mr.c, # input
            nSamp = 250, # desired sample number
            existing = existing, # existing samples
            iter = 100, # number of iterations
            cost = "dist2access", # cost parameter - name defined in calculate_distance()
            plot = TRUE) # plot

```

## `sample_balanced` {#balanced .unnumbered}

The `sample_balanced()` algorithm performs the balanced sampling methodology from the [`stratifyR / SamplingBigData`](http://www.antongrafstrom.se/balancedsampling/) packages.

```{r,warning=F,message=F} 
sample_balanced(mraster = mraster, # input
                nSamp = 200, # desired sample number
                plot = TRUE) # plot
```

```{r,warning=F,message=F} 
sample_balanced(mraster = mraster, # input
                nSamp = 100, # desired sample number
                algorithm = "lcube", # algorithm type
                access = access, # define access road network
                buff_inner = 50, # inner buffer - no samples within this distance from road
                buff_outer = 200) # outer buffer - no samples further than this distance from road
```

## `sample_ahels` {#ahels .unnumbered}

The `sample_ahels()` function performs the adapted Hypercube Evaluation of a Legacy Sample (ahels) algorithm using
`existing` sample data and an `mraster`. New samples are allocated based on quantile ratios between the `existing` sample and `mraster` covariate dataset.

This algorithm:

1. Determines the quantile distributions of `existing` samples and `mraster` covariates.

2. Determines quantiles where there is a disparity between samples and covariates.

3. Prioritizes sampling within those quantile to improve representation.

To use the function, the user will first specify the number of quantiles (`nQuant`) followed by either the `nSamp` (total number of desired samples to be added) or `threshold` (proportional representation between sample and covariate quantiles - default is 0.9) parameters. Its recommended to use `threshold` values at or below 0.9 as higher values can currently cause the algorithm to add samples repeatedly.

```{r,eval = FALSE}
sample_ahels(mraster = mraster[[1:3]], # input mraster - first 3 layers only
             existing = existing, # existing samples
             plot = TRUE) # plot
```

```{r,warning=F,message=F,echo=FALSE, results = FALSE}
s <- sample_ahels(mraster = mraster[[1:3]], # input mraster - first 3 layers only
             existing = existing, # existing samples
             plot = TRUE) # plot
```
```{r,echo=FALSE}
s
```

Notice that no `threshold`, `nSamp`, or `nQuant` were defined. Thats because defaults are `threshold = 0.9` and `nQuant = 10`.

The first matrix output shows the quantile ratios between sample and covariates. A value of 1.0 means that samples and covariates are equally represented. Values above 1.0 are over represented in the sample, less than 1 under represented in the sample.

```{r,eval = FALSE}
sample_ahels(mraster = mraster[[1:3]], # input mraster - first 3 layers only
             existing = existing, # existing samples
             nQuant = 20, # define 20 quantiles
             nSamp = 300, # total samples desired
             filename = tempfile(fileext = ".shp")) # write samples to disc
```

```{r,warning=F,message=F,echo=FALSE, results = FALSE}
s <- sample_ahels(mraster = mraster[[1:3]], # input mraster - first 3 layers only
             existing = existing, # existing samples
             nQuant = 20, # define 20 quantiles
             nSamp = 300, # total samples desired
             plot = TRUE,
             filename = tempfile(fileext = ".shp")) # write samples to disc

```

```{r,echo=FALSE}
s
```
Note that the total number of samples is 500. This is the total of existing samples (200) and the number defined by `nSamp = 300`.
