---
title: "sgsR"
output: rmarkdown::html_vignette
description: >
  Learn how to use sgsR.
vignette: >
  %\VignetteIndexEntry{sgsR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Algorithm structure {#str .unnumbered}

`sgsR` is primarily scripted using the `terra` package to handle raster processing and `sf` package for vector manipulation. Currently, there are 3 primary function verbs that this package uses:

*   `strat_*` - Stratify verb directs the functions to apply stratification algorithms to the input metrics raster `mraster` and produce stratified raster `sraster` as the output.

*   `sample_*` - Sample verb directs the functions to extract samples from `srasters`, which is produced from `strat_*` functions. Few algorithms (such as `sample_srs()`, `sample_balanced()`, `sample_systematic()`) are capable of using `mrasters` as the input because those algorithms do not depend on stratified inputs for sampling.  

*   `calculate_*` - Calculate verb directs the functions to perform calculations; values derived from these calculations are used in subsequent processing. Predefined sample analysis algorithms (such as `calculate_representation()`, `calculate_coobs()`) are included. 



## Example data {.unnumbered}
We demonstrate and provide examples of all the functions using the internal data from the package [specify the package]. Use the following code to load the internal data for `mraster` and road `access`. Follow along on your own device to explore different outputs and better comprehend the package functions.



### Metrics rasters - `mraster` {#mrast .unnumbered}
```{r,warning=F,message=F}
library(sgsR)
library(terra)
library(sf)

#--- Load mraster and access files ---#
r <- system.file("extdata", "wall_metrics.tif", package = "sgsR")

#--- load the mraster using the terra package ---#
mraster <- terra::rast(r)
```

### Road `access` data {#vect .unnumbered}
```{r,warning=F,message=F}
a <- system.file("extdata", "roads.shp", package = "sgsR")

#--- load the access vector using the sf package ---#
access <- sf::st_read(a)
```

```{r,warning=F,message=F}
terra::plot(mraster[[1]])
terra::plot(access, add = TRUE, col = "black")

```

From the plot we see the first band (`zmax`) of the `mraster` with the `access` vector overlaid. 

## Stratified rasters - `sraster` {#srast .unnumbered}

In this tutorial, I am going to demonstrate how to produce basic `sraster` and `existing` sample data, which will be used in the subsequent examples below. 

To produce `sraster`, we will be using the `strat_kmeans()` function. This function uses the kmeans algorithm to stratify the input `mraster`.

```{r,warning=F,message=F}
#--- apply kmeans algorithm to metrics raster ---#
sraster <- strat_kmeans(mraster = mraster, # use mraster as input for sampling
                        nStrata = 4, # algorithm will produce 4 strata
                        plot = TRUE) # algorithm will plot output

```

After `sraster` is produced, we will be applying the `sample_srs()` function. This function uses an algorithm which will randomly sample points within our `sraster` to produce `existing` sample dataset.

```{r,warning=F,message=F}
#--- set seed ---#
set.seed(2021)

#--- apply kmeans algorithm to metrics raster ---#
existing <- sample_srs(raster = mraster, # use mraster as input for sampling
                       nSamp = 200, # request 200 samples be taken
                       mindist = 100, # define that samples must be 100 m apart
                       plot = TRUE) # algorithm will plot output

```

Following all the previous steps will generate `mraster`, `access`, `sraster` and `existing` datasets. You should expect to see these generated datasets used for all the examples found in this document.

## `%>%` {#pipe .unnumbered}

The sgsR package leverages the `%>%` operator from the `magrittr` package. This allows us to "pipe" operations together to save in the amount of code needed to achieve an outcome. A simple example is demonstrated below.

```{r pipe, eval= FALSE}
#--- non piped ---#
sraster <- strat_kmeans(mraster = mraster, # use mraster as input for sampling
                        nStrata = 4, # algorithm will produce 4 strata
                        plot = TRUE) # algorithm will plot output

existing <- sample_srs(raster = sraster, # use mraster as input for sampling
                       nSamp = 200, # request 200 samples be taken
                       mindist = 100, # define that samples must be 100 m apart
                       plot = TRUE) # algorithm will plot output

extract_metrics(mraster = mraster,
                existing = existing)


#--- piped ---#
strat_kmeans(mraster = mraster, nStrata = 4) %>%
  sample_srs(., nSamp = 200, mindist = 100) %>%
  extract_metrics(mraster = mraster, existing = .)

```

