---
title: "stratification"
output: rmarkdown::html_vignette
description: >
  Learn how to use strat* functions.
vignette: >
  %\VignetteIndexEntry{stratification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r,warning=F,message=F,echo=FALSE,results=FALSE}
library(sgsR)
library(terra)
library(sf)

#--- Load mraster and access files ---#
r <- system.file("extdata", "mraster.tif", package = "sgsR")

#--- load the mraster using the terra package ---#
mraster <- terra::rast(r)

a <- system.file("extdata", "access.shp", package = "sgsR")

#--- load the access vector using the sf package ---#
access <- sf::st_read(a, quiet = TRUE)

#--- apply quantiles algorithm to metrics raster ---#
sraster <- strat_quantiles(mraster = mraster$zq90, # use mraster as input for sampling
                           nStrata = 4) # algorithm will produce 4 strata

#--- apply stratified sampling algorithm ---#
existing <- sample_strat(sraster = sraster, # use mraster as input for sampling
                         nSamp = 200, # request 200 samples be taken
                         mindist = 100) # define that samples must be 100 m apart

```

Fundamental to many structurally guided sampling approaches is the use of stratification methods that allow for more effective and representative sampling protocols. Below, I will identify the stratification methods included in `sgsR` and demonstrate some worked examples for each functions. 

Currently, there are 5 functions associated with the `strat` verb in the `sgsR` package:

*   `strat_kmeans()` - kmeans stratification

*   `strat_quantiles()`- quantile stratification

*   `strat_breaks()` - user-defined break stratification

*   `strat_poly()` - polygon attribute stratification

*   `strat_map()` - maps (combines) 2 stratifications

## `strat_kmeans` {#kmeans .unnumbered}

We will demonstrate a preliminary example of using the `strat_kmeans()` algorithm to prepare our `sraster` input data and provide additional examples below.

```{r,warning=F,message=F}
#--- perform stratification using k-means ---#
strat_kmeans(mraster = mraster, # input
             nStrata = 5) # algorithm will produce 4 strata

```

Notice that the code above did not plot anything. This is because `plot = FALSE` is set as the default for all functions in the `sgsR` package. Below, we will demonstrate examples where the `plot = TRUE` was set. 

```{r,warning=F,message=F}
strat_kmeans(mraster = mraster, # input
             nStrata = 10, # algorithm will produce 10 strata
             iter = 1000, # set minimum number of iterations to determine kmeans centers
             algorithm = "MacQueen", # use MacQueen algorithm
             plot = TRUE) # plot output
```

## `strat_quantiles` {.unnumbered}

The `strat_quantiles()` algorithm divides data into equally sized strata (`nStrata`). Similar to the `strat_breaks()` function, the (quantile) algorithm have the capability to perform stratification on a single `mraster`; users can also input a secondary `mraster` (`mraster2`) and specify the number associated with the desired strata (`nStrata2`). Note that the dual stratification output will result in a product of $nStrata * nStrata2$.

```{r,warning=F,message=F}
#--- perform quantiles stratification ---#
strat_quantiles(mraster = mraster$zq90,
                nStrata = 6,
                plot = TRUE)

#--- dual stratification - will produce 12 output strata ---#
strat_quantiles(mraster = mraster$zq90, 
                mraster2 = mraster$zsd,
                nStrata = 3, 
                nStrata2 = 4)
```

## `strat_breaks` {.unnumbered}

`strat_breaks()` function stratifies the data based on the breaks in covariates that the user defines. Single `mraster` can be defined or additional `mraster2` can also be defined. `breaks` and `breaks2` coincide with the user defined breaks for `mraster` and `mraster2` respectively. 

```{r,warning=F,message=F}
#--- perform stratification using user-defined breaks ---#

#--- define breaks for metric ---#
breaks <- c(seq(0,100,20))

breaks

#--- perform stratification using user-defined breaks ---#

values <- terra::values(mraster$zq90)

#--- define breaks for metric ---#
breaks2 <- c(5,10,15,20,25)

breaks2

```

Once the breaks are created, we can use them as input into the `strat_breaks` function using the `breaks` and `breaks2` parameters. 

```{r,warning=F,message=F}
#--- stratify on 1 metric only ---#
strat_breaks(mraster = mraster$pzabove2,
             breaks = breaks,
             plot = TRUE)
```

```{r,warning=F,message=F}
#--- stratify on 1 metric only ---#
strat_breaks(mraster = mraster$zq90,
             breaks = breaks2,
             plot = TRUE)
```   

## `strat_poly` {.unnumbered}

The `strat_poly()` algorithm stratifies the input based on spatial polygon attributes and features. Users may wish to stratify based on categorical or empirical variables that are not available through ALS data (e.g. species from forest inventory polygons). This function allows the user to define the attribute of interest as well as how the features within the attribute are to be grouped for stratification. 

Users define the input `poly` and its associated `attribute`. A `raster` layer must be defined to guide the spatial extent and resolution for the output stratification polygon. Based on the vector or list of `features`, stratification is applied and the polygon is rasterized into its appropriate strata. 

```{r}
#--- load in polygon coverage ---#
poly <- system.file("extdata", "inventory_polygons.shp", package = "sgsR")

fri <- sf::st_read(poly)
```

`attribute` column within the polygon must be defined and the `features` (within the `attribute`) must be specified to define the composition of the output strata. In our demonstration below, `attribute = "NUTRIENTS"` and features within `NUTRIENTS` (poor, rich, medium) will define the 3 desired output classes.   

```{r}
#--- stratify polygon coverage ---#
#--- specify polygon attribute to stratify ---#

attribute <- "NUTRIENTS"

#--- specify features within attribute & how they should be grouped ---#
#--- as a single vector ---#

features <- c("poor", "rich", "medium")

srasterpoly <- strat_poly(poly = fri, # input polygon
                          attribute = attribute, # attribute to stratify by
                          features = features, # features within attribute
                          raster = sraster, # raster to define extent and resolution for output
                          plot = TRUE) # plot output
```

`features` can be converted into amalgamated classes. In our demonstration below, `rich` and `medium` features are combined together, while `low` is left in isolation. The 2 vectors are specified into a list, which will result in the output of 2 classes (low & rich/medium).  

```{r}
#--- or as multiple lists ---#
g1 <- "poor"
g2 <- c("rich", "medium")

features <- list(g1, g2)

strat_poly(poly = fri,
           attribute = attribute,
           features = features,
           raster = sraster,
           plot = TRUE,
           details = TRUE)
```

You may have noticed that the `details` parameter was present in the demonstration code above. `details` parameter returns the output `outRaster`, the `$lookUp` table associated with the strata, and the polygon (`$poly`) is created to drive the stratification based on attributes and features specified by the users. 

## `strat_map` {.unnumbered}

There are situations where multiple levels of stratification may be sought-after. For instance, the user may wish to combine the output from `strat_poly()` (see the demonstration above) with 3 classes, with the 4 class kmeans stratification generated above in this vignette. 

The total number of classes of the output is multiplicative of the number of the strata. For example, if `sraster` has 3 strata and `sraster2` has 4 strata, then the output of `strat_map()` will be 12 strata in total. There may be occasions where stratum do not interact spatially, this will result in fewer output strata. 

```{r}
#--- map srasters ---#
strat_map(sraster = srasterpoly, # strat_poly 3 class stratification
          sraster2 = sraster, # strat_kmeans 4 class stratification
          plot = TRUE)

```

The convention for the numeric value of the output strata is the concatenation (merging) of `sraster` strata and `sraster2` strata. Check `$lookUP` for a clear depiction of this step. 

```{r}
strat_map(sraster = srasterpoly, # strat_poly 3 class stratification
          sraster2 = sraster, # strat_poly 3 class stratification
          stack = TRUE, # stack input and output strata into multi layer output raster
          details = TRUE, # provide additional details
          plot = TRUE) # plot output
```
