---
title: "stratification"
output: rmarkdown::html_vignette
description: >
  Learn how to use strat* functions.
vignette: >
  %\VignetteIndexEntry{stratification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r,warning=F,message=F,echo=FALSE,results=FALSE}
library(sgsR)
library(terra)
library(sf)

#--- Load mraster and access files ---#
r <- system.file("extdata", "wall_metrics.tif", package = "sgsR")

#--- load the mraster using the terra package ---#
mraster <- terra::rast(r)

a <- system.file("extdata", "roads.shp", package = "sgsR")

#--- load the access vector using the sf package ---#
access <- sf::st_read(a, quiet = TRUE)

#--- apply kmeans algorithm to metrics raster ---#
sraster <- strat_kmeans(mraster = mraster, # use mraster for stratification
                        nStrata = 4) # algorithm will plot output

#--- apply kmeans algorithm to metrics raster ---#
existing <- sample_srs(raster = sraster, # use sraster as input for sampling
                       nSamp = 200, # request 200 samples be taken
                       mindist = 100) # algorithm will plot output

```

Fundamental to many structurally guided sampling approaches is the use of stratification methods that allow for more effective and representative sampling protocols. Below I define the stratification methods that are included thus far in `sgsR` and provide some worked examples for each.

A total of 6 `strat` functions currently exist in the package:

*   `strat_kmeans()` - kmeans stratification

*   `strat_quantiles()`- quantile stratification

*   `strat_breaks()` - user-defined break stratification

*   `strat_poly()` - polygon attribute stratification

*   `strat_map()` - maps (combines) 2 stratification

## `strat_kmeans` {#kmeans .unnumbered}

We provide a preliminary example of the `strat_kmeans()` algorithm to prepare our `sraster` input data but here are a few more.

```{r,warning=F,message=F}
#--- perform stratification using k-means ---#
strat_kmeans(mraster = mraster, # input
             nStrata = 5) # algorithm will produce 4 strata

```

Notice that nothing plotted... That's because `plot = FALSE` is the default for all functions in `sgsR`. See examples below where `plot = TRUE`.

```{r,warning=F,message=F}
strat_kmeans(mraster = mraster, # input
             nStrata = 10, # algorithm will produce 10 strata
             iter = 1000, # set minimum number of iterations to determine kmeans centers
             algorithm = "MacQueen", # use MacQueen algorithm
             plot = TRUE) # plot output
```

```{r,warning=F,message=F}
strat_kmeans(mraster = mraster, # input
             nStrata = 5, # algorithm will produce 4 strata
             center = FALSE, # do not center data
             scale = FALSE, # do not scale data
             plot = TRUE, # plot output
             filename = tempfile(fileext = ".tif"), # write output sraster to file
             overwrite = TRUE) # overwrite file on disc if it exists

```

## `strat_quantiles` {.unnumbered}

The `strat_quantiles()` algorithm divides data into equally sized strata (`nStrata`). Similar to `strat_breaks()`, the algorithm allows for stratification on a single `mraster`, or the user can supply a secondary mraster (`mraster2`) and specify the associated number of desired strata (`nStrata2`). The dual stratification output will always result in the product of $nStrata * nStrata2$.

```{r,warning=F,message=F}
#--- perform quantiles stratification ---#
strat_quantiles(mraster = mraster$zq90,
                nStrata = 6,
                plot = TRUE)

#--- dual stratification - will produce 12 output strata ---#
strat_quantiles(mraster = mraster$zq90, 
                mraster2 = mraster$zsd,
                nStrata = 3, 
                nStrata2 = 4)
```
## `strat_breaks` {.unnumbered}

The `strat_breaks()` function stratifies data based on user-defined breaks in covariates. A single `metric` can be defined or an additional `metric2` can be supplied. `breaks` and `breaks2` correspond to the user-defined breaks for `metric` and `metric2` respectively.

```{r,warning=F,message=F}
#--- perform stratification using user-defined breaks ---#

#--- define breaks for metric ---#
breaks <- c(seq(0,100,20))

breaks

#--- perform stratification using user-defined breaks ---#

values <- terra::values(mraster$zq90)

#--- define breaks for metric ---#
breaks2 <- c(5,10,15,20,25)

breaks2

```

Once the breaks have been created we can input them into the function using the `breaks` and `breaks2` parameters.

```{r,warning=F,message=F}
#--- stratify on 1 metric only ---#

strat_breaks(mraster = mraster$pzabove2,
             breaks = breaks,
             plot = TRUE)
```

```{r,warning=F,message=F}
#--- stratify on 1 metric only ---#

strat_breaks(mraster = mraster$zq90,
             breaks = breaks2,
             plot = TRUE)
```   

## `strat_poly` {.unnumbered}

The `strat_poly()` algorithm stratifies based on spatial polygon attributes and features. A user may wish to stratify based on categorical or empirical variables that are not given by ALS data (e.g. species from forest inventory polygons). This method allows the user to define the attribute of interest as well as how features within attributes should be grouped for stratification.

The user defines the input `poly` and associated `attribute`. A `raster` layer must be provided as a guide for spatial extent and resolution for the output stratification polygon. Based on a vector or list of `features`, stratification is applied, and the polygon is rasterized into appropriate strata.

```{r}
#--- load in polygon coverage ---#
poly <- system.file("extdata", "inventory_polygons.shp", package = "sgsR")

fri <- sf::st_read(poly)
```

The `attribute` column must be defined, while the `features` within must be specified to define the number and composition of the output strata. In the case below the `attribute = "NUTRIENTS"` and features within `NUTRIENTS` (poor, rich, medium) will define the 3 desired output classes.

```{r}
#--- stratify polygon coverage ---#
#--- specify polygon attribute to stratify ---#

attribute <- "NUTRIENTS"

#--- specify features within attribute & how they should be grouped ---#
#--- as a single vector ---#

features <- c("poor", "rich", "medium")

srasterpoly <- strat_poly(poly = fri, # input polygon
                          attribute = attribute, # attribute to stratify by
                          features = features, # features within attribute
                          raster = sraster, # raster to define extent and resolution for output
                          plot = TRUE) # plot output
```

`features` can also be made to amalgamate classes. In the example below `rich` and `medium` features are combined while `low` is left alone. The 2 vectors are then added to a list, which outputs 2 classes (low & rich/medium).

```{r}
#--- or as multiple lists ---#
g1 <- "poor"
g2 <- c("rich", "medium")

features <- list(g1, g2)

strat_poly(poly = fri,
           attribute = attribute,
           features = features,
           raster = sraster,
           plot = TRUE,
           details = TRUE)
```

You will notice that the `details` parameter is present above. This returns the output `outRaster`, the `$lookUp` table associated with the strata, and the polygon (`$poly`) created to drive the stratification based on attribute and features provided by the users. 

## `strat_map` {.unnumbered}

There may be an instance where multiple levels of stratification are desired. For instance a user may want to combine the output from `strat_poly()` above with 3 classes, with the 4 class kmeans stratification `kmeans`.

The total number of classes will always be the multiplicative of the number of strata. i.e. if `sraster` has 3 strata and `sraster2` has 4 strata the output of `strat_map()` will be 12 strata total.

```{r}
#--- map srasters ---#
strat_map(sraster = srasterpoly, # strat_poly 3 class stratification
          sraster2 = sraster, # strat_kmeans 4 class stratification
          plot = TRUE)

```

The convention for the numeric value of output strata is the concatenation (merging) of sraster strata and sraster2 strata. See `$lookUp` for a clear depiction of this.

```{r}
strat_map(sraster = srasterpoly, # strat_poly 3 class stratification
          sraster2 = sraster, # strat_poly 3 class stratification
          stack = TRUE, # stack input and oputput strata into multi layer ouput raster
          details = TRUE, # provide additional details
          plot = TRUE) # plot output
```